---
title: "Point Cloud Tree Analysis"
author: "Wade Sedgwick"
date: "2025-03-31"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
# Load libraries
library(lidR)
library(raster)
library(sp)
library(rgdal)
library(rgl)
library(sf)
library(dplyr)
library(ggplot2)
library(ForestTools)
library(TreeLS)
library(dbscan)
library(viridis)
library(mapview)
library(sf)
```

### Step 1-4: load point cloud, classify ground, generate DTM & normalize height
```{r}
las_file <- '/Volumes/ML_data/carbon/2025 04 LLP surveys/orthomosaics and point clouds/point clouds/millpond big pine 100m point_cloud/20250410 millpond big pine_group1_densified_point_cloud_part_1.las'

# 1. read las file
las <- readLAS(las_file)

# 2. classify ground using CSF (cloth simulation)
las <- classify_ground(las, csf()) # classifies ground to value of 2

# 3. generate DTM
dtm <- rasterize_terrain(las, res = 0.25, algorithm = tin())
# tin (Delaunay triangulation) performs linear interpolation within each triangle

# 4. normalize height (create CHM)
las_norm <- normalize_height(las, dtm)
```

```{r}
# viewing area on a satellite map
bbox <- st_bbox(las)
poly <- st_as_sfc(bbox)
st_crs(poly) <- 32617
mapview(poly, map.types = "Esri.WorldImagery", col.regions = "red", alpha.regions = 0.3)
```

Algorithms for classifying ground:
- noise classification alg.
  - sor(statistical outlier removal) - general outliers
  - ivf(isolated voxel filter) - isolated/noisy points
- ground classification alg.
  - pmf(progressive morphological filter) - flat to rolling terrain
  - csf(cloth simulation filter) - complex/mountainous terrain
  - mcc(multiscale curvature classification) - natural & variable
    - use mcc() when pmf() struggles with over-filtering, or csf() is too smooth

### Step 5: Remove Outliers + Ground Points from Normalized Data
```{r}
# print
table(las_norm@data$Classification)

# remove ground points
las_nonground <- filter_poi(las_norm, Classification != 2)

# get max canopy height (98th percentile)
canopy_max <- quantile(las_nonground$Z, probs = .98, na.rm = TRUE)

# keep relevant above-ground points
las_filtered <- filter_poi(las_nonground, Z > 0 & Z < (canopy_max + 10))

# classify noise using SOR algorithm
las_filtered <- classify_noise(las_filtered, algorithm = sor(k = 8, m = 2)) # classifies noise to value of 18
# uses Statistical Outliers Removal method - computes mean distance to KNN

# remove noise
table(las_filtered@data$Classification)
las_filtered <- filter_poi(las_filtered, Classification != 18)
summary(las_filtered$Z)

plot(las_filtered, color = "Classification")
# plot(las_norm, color = "Classification")
rglwidget()
```


```{r}
# compare histogram of heights before & after filtering
par(mfrow = c(1, 2))
hist(las_norm$Z, breaks = 50, main = "Before Filtering")
hist(las_filtered$Z, breaks = 50, main = "After Filtering")

plot(las_filtered)
rglwidget()
```


### Step 6: CHM & Tree Detection
```{r}
# generate DSM
dsm <- rasterize_canopy(las_filtered, res = 0.25, algorithm = p2r()) # points to raster method

# DSM = CHM
chm <- dsm

# storing na values as 0
# chm[is.na(chm[])] <- 0

# clean CHM - removes outer rows & columns that are NA
# chm_clean <- terra::trim(chm)

summary(chm$Z)

# variable window size - scales proportionally to tree height
vwf_safe <- function(x) pmax(0.2*x, 1)

# local max filter - only identifies trees and outputs height and geometry
trees <- locate_trees(chm, lmf(ws = vwf_safe, hmin = 1.37))

# extract tree heights
tree_sf <- st_as_sf(trees, coords = c("x", "y"), crs = crs(chm))
tree_sf$Z <- trees$height

summary(trees$Z)
```

### Exploratory Viz to view growth stages & individual trees
Viz: Viewing Longleaf Pine Growth Stages (crowns missing so not an accurate representation)
```{r}
# assign growth stages - crowns missing so not an accurate representation
summary_table <- trees %>%
  mutate(
    Growth_Stage = cut(
      Z,
      breaks = c(-Inf, 0.5, 3, 10, 20, Inf),
      labels = c("Grass Stage (<0.5m)", "Bottlebrush (0.5-3m)", "Sapling (3-10m)",
                 "Pole (10-20m)", "Mature (>20m)"),
      right = FALSE
    )
  ) %>%
  st_drop_geometry() %>% 
  count(Growth_Stage, name = "Count", .drop = FALSE)

print(summary_table)
```

Viz: visualize individual tree point cloud
```{r}
# display a point cloud of a tree - some trees will require wider area around stem
# extract XYZ and bind to the original sf object
tree_sf <- tree_sf %>% 
  bind_cols(st_coordinates(tree_sf) %>% as_tibble())

# filter Z > 10
tree_slice <- tree_sf %>% 
  filter(Z > 10) %>% 
  slice(2)

coords <- st_coordinates(tree_10m)


# crop a 3 m x 3 m area around the stem
las_crop <- clip_rectangle(las_filtered,
                             xleft = coords[1] - 1.5, 
                             xright = coords[1] + 1.5,
                             ybottom = coords[2] - 1.5, 
                             ytop = coords[2] + 1.5)

# filter Z/height to btwn 1.15 and 1.5 m
las_crop_dbh <- filter_poi(las_crop, Z >= 1.15 & Z <= 1.5)

# plotting tree
plot(las_crop, color = "Z", pal = viridis::viridis(100))
rglwidget()
```


```{r}
# plotting DBH slice
plot(las_crop_dbh, color = "Z", pal = viridis::viridis(100))
rglwidget()
```


```{r}
# plotting individual tree on a satellite map
bbox <- st_bbox(las_crop)
poly <- st_as_sfc(bbox)
st_crs(poly) <- 32617
mapview(poly, map.types = "Esri.WorldImagery", col.regions = "red", alpha.regions = 0.3)
```



### Step 7-8: Extracts dbh slice for each tree location, filters DBH slices based on params, and fits circle to XY coords
```{r}
# extract dbh slice per known tree location of point cloud
tree_coords <- st_coordinates(tree_sf)
tree_data <- cbind(tree_sf, tree_coords) %>%
  st_drop_geometry()

# parameters
dbh_height <- 1.37       # breast height
dbh_thickness <- 0.2     # +/- 10cm slice
dbh_radius <- 0.4        # 40cm horizontal buffer around tree center

all_slices <- list()

# Step 8: filters DBH slices based on parameters
for (i in seq_len(nrow(tree_data))) {
  x0 <- tree_data$X[i]
  y0 <- tree_data$Y[i]
  id <- tree_data$treeID[i]
  
  # subset all points within 40 cm horizontally
  local_slice <- clip_circle(las_filtered,
                             xcenter = x0, ycenter = y0, radius = dbh_radius)
  
  # filter vertically near DBH height
  local_slice <- filter_poi(local_slice, Z >= (dbh_height - dbh_thickness / 2) & Z <= (dbh_height + dbh_thickness / 2))
  
  # stores slice if it has enough points
  if (npoints(local_slice) > 10) {
    slice_df <- as.data.frame(local_slice@data)
    slice_df$treeID <- id
    all_slices[[length(all_slices) + 1]] <- slice_df
  }
}

dbh_slices_df <- bind_rows(all_slices)


# get the first 10 treeIDs
n <- 10
valid_ids <- dbh_slices_df %>%
  distinct(treeID) %>%
  arrange(treeID) %>%
  slice(1:n) %>%
  pull(treeID)

# plot only those trees to visualize roundness
ggplot(filter(dbh_slices_df, treeID %in% valid_ids), aes(x = X, y = Y)) +
  geom_point(aes(color = rgb(R / 65535, G / 65535, B / 65535)), size = 0.6) +
  scale_color_identity() +
  facet_wrap(~ treeID, scales = "free") +
  theme_minimal() +
  labs(title = paste("DBH Slices (with RGB values) – First", n, "Trees"),
       x = "X", y = "Y")


all_dbhs <- list()

# Step 8: extracts DBH estimates using RANSAC circle-fitting method and fits circle to XY coords - uses TreeLS library
for (tree_id in unique(dbh_slices_df$treeID)) {
  pts <- dbh_slices_df %>% filter(treeID == tree_id)
  las_tree <- LAS(pts, header = las_norm@header)
  las_tree <- add_attribute(las_tree, rep(1L, nrow(pts)), "Stem")

  result <- tryCatch({
    tlsInventory(las = las_tree, dh = dbh_height, dw = dbh_radius, hp = 1,
                 d_method = shapeFit(shape = "circle", algorithm = "ransac")) # random sample consensus - resamples until subset only has inliers
  }, error = function(e) NULL)
  
  if (!is.null(result)) {
    df <- as.data.frame(result)
    df$treeID <- tree_id
    all_dbhs[[length(all_dbhs)+1]] <- df
  }
}

dbh_result_df <- bind_rows(all_dbhs)
dbh_result_df$DBH_cm <- round(dbh_result_df$Radius * 2 * 100, 1)

# visualize
head(dbh_result_df, 2)
head(dbh_slices_df, 2)
```

### Exploratory viz to show DBH outliers
```{r}
# exploratory - heights compared to DBH to see how far off the heights are
merged_df <- dbh_result_df %>%
  left_join(tree_sf %>% st_drop_geometry() %>% select(treeID, Z), by = "treeID") %>% 
  mutate(Z_pred_m = (1.3 + exp(0.8936 + 0.8096 * log(DBH_cm))) / 3.28084,
         Z_error = Z - Z_pred_m)

```

#### Filtering trees with DBH greater than than 90 cm to view outliers

```{r}

# showing trees with DBH > 90 cm
# Filter trees with DBH > 90 cm
big_trees <- dbh_result_df %>% filter(DBH_cm > 90)

# Convert to sf points
big_trees_sf <- st_as_sf(big_trees, coords = c("X", "Y"), crs = 32617)

# Create buffer around each tree (e.g., 1 m radius)
big_tree_buffers <- st_buffer(big_trees_sf, dist = 1)

# Clip all the trees using the union of the buffers
big_tree_area <- st_union(big_tree_buffers)

# Clip LAS points within those areas
big_tree_las <- clip_roi(las_filtered, big_tree_area)

plot(big_tree_las, color = "Z", pal = viridis::viridis(100))
rglwidget()

```

```{r}
# showing trees with DBH greater than 90 cm
# Choose specific rows by index (e.g., rows 1 to 3)
idx <- seq(1, nrow(big_trees))



# Buffer and plot each tree
for (i in idx) {
  pt <- st_point(c(big_trees$X[i], big_trees$Y[i]))
  sf_pt <- st_sfc(pt, crs = 32617) |> st_buffer(dist = 1)
  las_tree <- clip_roi(las_filtered, sf_pt)

  if (npoints(las_tree) < 10) next

  plot(las_tree, color = "Z", pal = viridis::viridis(100))
  print(rglwidget())
  readline("Press [Enter] for next tree...")
}

```

```{r}
# viewing bottom 2 meters of trees with DBH greater than 90 cm with horizontal plane at 1.37 m
idx <- seq(1, 6)

for (i in idx) {
  # Create buffered point around tree location
  pt <- st_point(c(big_trees$X[i], big_trees$Y[i]))
  sf_pt <- st_sfc(pt, crs = 32617) |> st_buffer(dist = 1)
  
  # Clip LAS and filter to bottom 5 meters
  las_tree <- clip_roi(las_filtered, sf_pt)
  las_bottom <- filter_poi(las_tree, Z <= 2)

  if (npoints(las_bottom) < 10) next  # Skip if too few points

  # Plot just the bottom 5 meters
  plot(las_bottom, color = "Intensity", pal = rep("gray", 100))
  planes3d(a = 0, b = 0, c = 1, d = -1.37, col = "red", alpha = 0.9)
  print(rglwidget())
  readline("Press [Enter] for next tree...")
}

```

#### Plotting DBH slices to identify outliers

```{r}
# plotting slices all centered on (0,0)
dbh_slices_centered <- dbh_slices_df %>%
  inner_join(dbh_result_df %>% select(treeID, X_center = X, Y_center = Y), by = "treeID") %>%
  mutate(
    X_rel = X - X_center,
    Y_rel = Y - Y_center
  )

# plotting DBH slices centered, colors representing true RGB color values
ggplot(dbh_slices_centered, aes(x = X_rel, y = Y_rel)) +
  geom_point(aes(color = rgb(R / 65535, G / 65535, B / 65535)), size = 0.6, alpha = 0.8) +
  scale_color_identity() +
  coord_fixed() +
  theme_minimal() +
  labs(title = "All DBH Slices Centered at (0, 0)", x = "X (centered)", y = "Y (centered)")

library(ggforce)

# plotting centered slices with fitted circles
ggplot(dbh_slices_centered, aes(x = X_rel, y = Y_rel)) +
  geom_point(aes(color = rgb(R / 65535, G / 65535, B / 65535)), size = 0.6, alpha = 0.8) +
  geom_circle(data = dbh_result_df, aes(x0 = 0, y0 = 0, r = Radius), 
              color = "red", inherit.aes = FALSE, alpha = 0.5) +
  scale_color_identity() +
  coord_fixed() +
  theme_minimal() +
  labs(title = "All Centered DBH Slices with Fitted Circles", x = "Centered X", y = "Centered Y")

# plot showing the different trees to reveal outliers
ggplot(dbh_slices_centered, aes(x = X_rel, y = Y_rel, color = factor(treeID))) +
  geom_point(size = 0.6, alpha = 0.8) +
  coord_fixed() +
  theme_minimal() +
  labs(title = "DBH Slices Centered by Tree(n=150)", x = "X (meters)", y = "Y (meters)", color = "Tree ID")

## filter to remove green point cloud points - some trees will have green vegetation at the 1.37 m height

```

### Step 10: Filtering DBH less than 90 cm and using biomass alometric equation from Chojnacky et al. 2014 for southern pines (Done with outliers)

```{r}
# Filter DBHs < 90 cm
small_trees <- dbh_result_df %>% filter(DBH_cm < 90)

# Apply allometric equation
small_trees <- small_trees %>%
  mutate(
    AGB_kg = 0.0509 * (DBH_cm ^ 2.5433),
    Carbon_kg = 0.5 * AGB_kg,
    AGB_min = AGB_kg * 0.8,
    AGB_max = AGB_kg * 1.2,
    Carbon_min = Carbon_kg * 0.8,
    Carbon_max = Carbon_kg * 1.2
  )

small_trees %>% select(treeID, DBH_cm, AGB_kg, Carbon_kg) %>% head()

carbon_summary <- data.frame(
  Stat = c("Estimated", "Min", "Max"),
  Carbon_tons = round(c(
    sum(small_trees$Carbon_kg) / 1000,
    sum(small_trees$Carbon_min) / 1000,
    sum(small_trees$Carbon_max) / 1000
  ), 2)
)

print(carbon_summary)
# only using DBH has ~20-30% accuracy, whereas with height, it has ~10-15% accuracy

```

```{r}
# viewing area on satellite map
bbox <- st_bbox(las)
poly <- st_as_sfc(bbox)
st_crs(poly) <- 32617
mapview(poly, map.types = "Esri.WorldImagery", col.regions = "red", alpha.regions = 0.3)
```



### Visualize a point cloud for a single tree - uses convex hull method vs RANSAC using tlsInventory()
```{r}
# visualizing one tree point cloud
tree_pt <- dplyr::filter(tree_sf, treeID == 16)
buffer_radius <- 0.5
tree_buf <- st_buffer(tree_pt, dist = buffer_radius)

# Clip the LAS file to the buffer
tree_las <- clip_roi(las, tree_buf)

# normalize
las_norm_pt <- normalize_height(tree_las, dtm)  # dtm is a raster or Terrain Model

# slice 1.15 to 1.5 meters
dbh_slice_pt <- filter_poi(las_norm_pt, Z >= 1.15 & Z <= 1.5)

# extract x and y from dbh slice
xyz <- dbh_slice_pt@data[, c("X", "Y")]

# calculating DBH for the chosen tree
if (nrow(xyz) >= 3) {
  # Fit a minimum enclosing circle using convex hull method
  ch <- chull(xyz)
  ch <- c(ch, ch[1])
  hull_pts <- xyz[ch, ]

  # Approximate DBH as diameter of bounding circle
  centroid <- colMeans(hull_pts)
  distances <- sqrt((hull_pts[,1] - centroid[1])^2 + (hull_pts[,2] - centroid[2])^2)
  dbh_est <- 2 * max(distances)  # in meters
  dbh_cm <- dbh_est * 100        # convert to cm

  cat("Estimated DBH:", round(dbh_cm, 1), "cm\n")
} else {
  cat("Not enough points in DBH slice.\n")
}

plot(xyz)
plot(dbh_slice_pt)
rglwidget()

```

```{r}
# plotting individual tree on a satellite map
# extent as a bounding box
bbox <- st_bbox(dbh_slice_pt)

# Convert to an sf polygon
poly <- st_as_sfc(bbox)

# Set coordinate ref system
st_crs(poly) <- 32617

mapview(poly, map.types = "Esri.WorldImagery", col.regions = "red", alpha.regions = 0.3)
```


